Lossy compression of DNA sequencing quality data

### 2015-04-13 manuscript/q-compression.tex
Added outline of alignment results.

Generally, increasing the regression model polynomial order from 0 to
5 results in more mapped reads with the most marked change from order
1 to order 2. The best compression approach is the one that has
highest proportion of reads shared between the uncompressed and
decompressed files and least number of reads that are unique in both
uncompressed and decompressed files. In fragment sample, on average,
99.03\% of reads were mapped using different compression approaches,
within which 99\% are shared with the uncompressed alignments and
0.02\% unique mapped reads in the decompressed file. The average
missed alignments are 0.99\%. Compression with profile 256 approach
was shown to be the best performing approach in the fragment
decompressed sample, as a proportional of 99.94\% reads are mapped
with respect to the uncompressed mapped reads within which 99.91\% are
common and only 0.08\% unique mapped reads in the decompressed
file. In short-run sample, the average number of mapped reads is 99.14
\% within which 98.73\% are common and 0.41\% unique mapped reads in
the decompressed file.The average missed alignments are
1.2\%. Regression with the 5th polynomial order captured the highest
mapped reads of 100\%, within which 99.66\% are common and only 0.33\%
unique mapped reads in the decompressed file.

\begin{table*}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the number
  of reads that are kept or discarded after applying compression with
  respect to the original, uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
  \cline{2-12}
& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
\cline{2-12}
 & sum & 1025434 &   0 &   0 & 1025434 & 872917 & 152517 & 729824 & 295610 & 880724 & 144710 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
\cline{2-12}
&  sum & 882375 & 143059 & 888940 & 136494 & 886623 & 138811 & 891944 & 133490 & 894966 & 130468 \\
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
 & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
\cline{2-12}
&  sum & 897200 & 128234 & 892136 & 133298 & 892537 & 132897 & 892820 & 132614 & 889118 & 136316 \\ 
\end{tabular}
\end{table*}

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\

>>>>

We evaluate the compression methods using two approaches.  In the first approach, we order the compression  methods based on how similar the alignment results are to the uncompressed sequences, i.e., the amount of reads aligned by both the uncompressed and compressed methods minus the amount of reads uniquely aligned by the uncompressed and compression method.  In the second approach, we order the compression methods by total proportion of aligned reads.

The best compression method in terms of similarity with the uncompressed reads is QualComp with rate 100 bits/read, followed by QualComp with rate 30 bits/read, 256-, and 128-profile encoding, then 2-bin encoding, 64-profile encoding, QualComp with rate 10 bits/read, 7th-degree polynomial regression, QualComp with rate 6 bits/read, and finally, 5th-degree through 0-degree polynomial regression.

Ranking the compression methods by overall alignment rate produces an identical ordering as above.  Aside from 0-degree polynomial regression (83.1\%), all other compression methods have an alignment rate between 87\% and 86.1\%.

Most of the compression methods did not vary greatly in terms of the number of reads that were mapped \emph{only} by the compression method; however, there is a sizable difference in the amount of reads that are originally mapped, but unmapped by the compressed methods.
QualComp with rate 100 bits/read results in the fewest missing original read alignments (159).
Increasing the regression model polynomial degree results in a decreasing amount of reads that are originally mapped, but unmapped by the regression model (40,931 to 1,134 reads for 0-degree and 7th-degree, respectively).
There is no such trend for reads that are mapped only by the regression model.

Setting all bases as minimum quality results in the highest fraction of mapped reads 88.1\%.
Conversely, setting all bases as maximum quality results in the lowest fraction of mapped reads 72.8\%.

\begin{table*}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the number
  of reads that are kept or discarded after applying compression with
  respect to the original, uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
  \cline{2-12}
& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
\cline{2-12}
 & fraction & 1.000 & 0.000 & 0.000 & 1.000 & 0.851 & 0.149 & 0.712 & 0.288 & 0.859 & 0.141 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
\cline{2-12}
& fraction & 0.860 & 0.140 & 0.867 & 0.133 & 0.865 & 0.135 & 0.870 & 0.130 & 0.873 & 0.127 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
 & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
\cline{2-12}
& fraction & 0.875 & 0.125 & 0.870 & 0.130 & 0.870 & 0.130 & 0.871 & 0.129 & 0.867 & 0.133 \\ 
\end{tabular}
\end{table*}

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
%& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
& fraction & 0.728 & 0.272 & 0.881 & 0.119 & 0.870 & 0.130 & 0.831 & 0.169 & 0.862 & 0.138 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
%& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
& fraction & 0.868 & 0.132 & 0.869 & 0.131 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\ 
\end{tabular}

@@

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
& &  mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
%& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\ 
& fraction & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 & 0.870 & 0.130 \\
